---
title: "Figure 1"
author: "Joao Carlos Gomes-Neto"
date: "8/2/2021"
output: html_document
---

When beginning running this script, you can remove the # that comes before the install.packages() function, but when running it the second time around, and subsequently, comment the function out using #. That way you avoid creating issues with dependencies, and different versions of the program. 

```{r, include = TRUE}
# install tidyverse 
# install.packages("tidyverse")
# install skimr 
# install.packages("skimr")
# install vegan 
# install.packages("vegan")
# install forcats
# install.packages("forcats")
# install naniar
# install.packages("naniar")
# install ggpubr
# install.packages("ggpubr")
# install ggrepel
# install.packages("ggrepel")
# install reshape2
# install.packages("reshape2")
# install RColorBrewer
# install.packages("RColorBrewer")
# install ggtree 
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("ggtree")
# installation of ggtree will prompt a question about installation - answer is "a" to install/update all dependencies 
```

How to activate packages prior to utilization.

```{r, include = TRUE}
# Load previously installed packages
library(tidyverse)
library(skimr)
library(vegan)
library(forcats)
library(naniar)
library(ggtree)
library(ggpubr)
library(ggrepel)
library(reshape2)
library(RColorBrewer)
```

Enter and quality control all genotypic data including: serovar-predictions (generated by SISTR), BAPS level 1 (generated by fastbaps), ST lineages (generated by mlst), 
and cgMLST variants (generated by SISTR). All input files were generated by the describe programs which are part of the computational platform called ProkEvo.

```{r, include = TRUE}
# enter the BAPS data 
baps <- read_csv('fastbaps_partition_baps_prior_l6.csv')
# check the first six rows of the dataset
head(baps)
# changing the first two column names because for hierarchical analysis we only use BAPS1
colnames(baps)[1:2] <- c("id", "baps1") 
# check the first six rows of the datset again to see the change in column names
head(baps)
# select columns id and baps_1
baps1 <- baps %>% 
              select(id, baps1)
# check the first six rows of the dataset
head(baps1)
# quality control baps1 data 
skim(baps1)
# using a plotting strategy to check for missing values
vis_miss(baps1)
# rename BAPS level 1 sub-groups or haplotypes 
baps1$baps_1 <- ifelse(baps1$baps1 == 1, "BAPS1 sub-group 1", # the ifelse functions tests for different conditions prior to assigning groups to one category or another 
                   ifelse(baps1$baps1 == 2, "BAPS1 sub-group 2", 
                          ifelse(baps1$baps1 == 3, "BAPS1 sub-group 3", 
                                 ifelse(baps1$baps1 == 4, "BAPS1 sub-group 4", 
                                        ifelse(baps1$baps1 == 5, "BAPS1 sub-group 5", 
                                               ifelse(baps1$baps1 == 6, "BAPS1 sub-group 6", 
                                                       ifelse(baps1$baps1 == 7, "BAPS1 sub-group 7", 
                                                               ifelse(baps1$baps1 == 8, "BAPS1 sub-group 8", "BAPS1 sub-group 9"))))))))
##########################################--------############################################
#########################################--------#############################################
# enter MLST results
mlst <- read_csv('salmonellast_output.csv')
# check the first six rows of the dataset
head(mlst)
# generate the id column by deriving it from the FILE column
mlst$id <- sapply(strsplit(as.character(mlst$FILE),'_'), "[", 1)
# check the first six rows of the dataset
head(mlst)
# select the id and ST columns 
mlst1 <- mlst %>%
          select(id, ST) # select id and ST columns 
# use the table() function to detect extraneous characters such as "-" or "?" in the data - those are ST misclassifications
# check for the presence of ST misclassification in the ST column 
table(mlst1$ST)
# check for missing values
skim(mlst1)
# mutate the "-" character to NA (missing value)
mlst1 <- mlst1 %>%
            mutate_all(na_if, "-") # fill all - with NAs
# check the first six rows of the dataset
head(mlst1)
# quality control baps1 data 
skim(mlst1)
# using a plotting strategy to check for missing values
vis_miss(mlst1)
# At this stage missing values are not removed or dealt with until datasets are merged 
##########################################--------############################################
#########################################--------#############################################
# Enter SISTR results
sistr <- read_csv('sistr_output.csv')
# check the first six rows of the dataset
head(sistr)
# generate the id column by deriving it from the genome column
sistr$id <- sapply(strsplit(as.character(sistr$genome),'_'), "[", 1)
# select the id and cgmlst_ST columns 
sistr1 <- sistr %>%
            select(id, serovar_cgmlst, cgmlst_ST) # select id, serovar_cgmlst, and cgmlst_ST columns 
# check the first six rows of the dataset
head(sistr1)
# quality control baps1 data 
skim(sistr1)
# using a plotting strategy to check for missing values
vis_miss(sistr1)
# At this stage missing values are not removed or dealt with until datasets are merged 
##########################################--------############################################
#########################################--------#############################################
# combine baps1, mlst1, and sistr1 datasets
d1 <- left_join(baps1, mlst1, on = "id") # merge datasets on identical ids 
d2 <- left_join(d1, sistr1, on = "id") # merge datasets on identical ids 
# check data dimensionality 
dim(d2)
# check the first six rows of the dataset
head(d2)
# quality control baps1 data 
skim(d2)
# using a plotting strategy to check for missing values
vis_miss(d2)
# At this stage missing values are not removed or dealt with until datasets are merged 
##########################################--------############################################
#########################################--------#############################################
# group all genomes not classified as Newport as "Other serovars"
# create a new column called serovar that contains the binary classification to group all SISTR-based misclassified genomes
d2$serovar <- ifelse(d2$serovar_cgmlst == "Newport", "Newport",
                            "Other serovars") # classify serovar_cgmlst into Newport or Others 
# check the first six rows of the dataset
head(d2)
# check the composition of the serovar column 
table(d2$serovar)
##########################################--------############################################
#########################################--------#############################################
# no transformation is needed for the baps1 dataset - no groupings or aggregations are needed 
##########################################--------############################################
#########################################--------#############################################
# check ST distribution to focus on major STs for all subsequent analyses by grouping by ST and counting
st_dist <- d2 %>%
  group_by(ST) %>% # group by the ST column
  count() %>% # count the number of observations 
  arrange(desc(n)) # arrange the counts in decreasing order 
st_dist
# based on the frequency analysis, the following STs were not aggregrated: ST118, ST45, ST5, ST132, ST31, ST350, and ST46
# create a new st column
d2$st <- ifelse(d2$ST == 5, "ST5", # create a new ST column for which minor STs are aggregated as Others 
                   ifelse(d2$ST == 31, "ST31", 
                          ifelse(d2$ST == 45, "ST45", 
                                 ifelse(d2$ST == 46, "ST46", 
                                        ifelse(d2$ST == 118, "ST118", 
                                               ifelse(d2$ST == 132, "ST132", 
                          ifelse(d2$ST == 350, "ST350", "Other STs")))))))
# check the first six rows of the dataset
head(d2)
##########################################--------############################################
#########################################--------#############################################
# check cgMLST distribution by ST to focus on major cgMLSTs for all subsequent analyses by grouping by STs and cgMLSTs and counting
cgmlst_dist <- d2 %>%
  group_by(st, cgmlst_ST) %>% # group by st and cgmlst_ST
  count() %>% # count the number of observations 
  arrange(desc(n)) # arrange in descending order 
cgmlst_dist
# for the purposes of this analysis only the top four most frequent cgMLSTs were selected and respectively renamed
d2 <- mutate(d2, cgmlst = ifelse(cgmlst_ST %in% 1468400426, "cgMLST 1468400426", # create a new cgMLST column while aggregating minor cgMLST variants 
                                     ifelse(cgmlst_ST %in% 88443731, "cgMLST 88443731", 
                                            ifelse(cgmlst_ST %in% 1271156802, "cgMLST 1271156802",
                                                 ifelse(cgmlst_ST %in% 3336043520, "cgMLST 3336043520",  
                                            "Other cgMLSTs")))))
# check the first six rows of the dataset
head(d2)
##########################################--------############################################
#########################################--------#############################################
# select only needed columns for all subsequent analyses 
d3 <- d2 %>% 
          select(id, serovar, baps_1, st, cgmlst) # select columns of interest which are described within parenthesis 
# check data dimensionality to make sure it matches that of d2 
dim(d3)
# check the first six rows of the dataset
head(d3)
# quality control baps1 data 
skim(d3)
# using a plotting strategy to check for missing values
vis_miss(d3)
# make sure there is no NA in the dataset
sum(is.na(d3))
# replace NA in the ST column
d3 <- d3 %>% mutate(st = replace_na(st, "Other STs")) 
# make sure there is no NA in the dataset
sum(is.na(d3))
# check the distribution of serovars, baps1, st, and cgmlst classifications
table(d3$serovar)
table(d3$baps_1)
table(d3$st)
table(d3$cgmlst)
# check the first six rows in the dataset 
head(d3)
# check the last six rows in the dataset 
tail(d3)
# check for missing values
vis_miss(d3)
# check the data characteristics 
str(d3)
```

Figure 1. Visualize the phylogeny-based hierarchical distribution of genotypes.

At the center of the figure is the core-genome phylogeny. Genotypic information is plotted onto it in the following order:
Serovar (innermost)
BAPS1
ST
cgMLST (outermost)

```{r, include = TRUE}
# enter the phylogeny data
tree <- read.tree("newport_phylogeny.tree")
# recall the metadata combining all hierarchical genotypes 
# here I assign d3 to d4 just to keep the formatted metadata read for other steps, in case I need to do 
# any transformation with the data to plot with the phylogenetic tree
d4 <- d3
# to plot with the phylogeny using ggtree we need to make the id column into index first 
d4 <- column_to_rownames(d4, var = "id") 
# create the tree
# adjusting the parameters to make the tree visible or however you wish requires some trial and error 
tree_plot <- ggtree(tree, layout = "circular") + xlim(-250, NA)
# plot figure 1 - color scheme for each layer of the plot should be chosen based on the user preferences 
figure_1 <- gheatmap(tree_plot, d4, offset=.0, width=20, colnames = FALSE) +  # visualize the tree with metadata 
  scale_fill_manual(values = c("coral", "darkblue",
                               "cornflowerblue", "coral", "purple", "red", "brown", "darkseagreen3", "darkblue", "darkgreen", "yellow",
                               "orange", "darkblue", "purple", "darkgreen", "darkred", "steelblue", "black", "coral",
                               "black", "darkgreen", "purple", "darkblue", "gray"),
                    breaks = c("Newport", "Other serovars",
                               "BAPS1 sub-group 1", "BAPS1 sub-group 2", "BAPS1 sub-group 3", "BAPS1 sub-group 4", 
                               "BAPS1 sub-group 5", "BAPS1 sub-group 6", "BAPS1 sub-group 7", "BAPS1 sub-group 8", 
                               "BAPS1 sub-group 9",
                               "ST5", "ST31", "ST45", "ST46", "ST118", "ST132", "ST350", "Other STs",
                               "cgMLST 1468400426", "cgMLST 1271156802", "cgMLST 3336043520", "cgMLST 88443731", "Other cgMLSTs"),
                    name="Serovar (innermost) -> BAPS1 -> ST -> cgMLST (outermost)") +  # add colors and labels for the scale
    ggtitle(expression(bold(paste("Hierarchical population structure - ", bolditalic(S),". Newport")))) +  # add title for the figure
  theme(plot.title = element_text(size = 40, face = "bold"), 
        legend.title=element_text(size=24, face = "bold"), 
    legend.text=element_text(size=22))  # customize figure's title, legend, font
figure_1
```
